Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-29 supplement.Rmd
CL comments addressed

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.  For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.

Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.  Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.  Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.  Data for *K. brevis* were obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).  Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St. Petersberg) impacted by the event. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).

Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.  Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17]. Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.  This rapid survey design was created by the "Eyes on Seagrass" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant. Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.  Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.  Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.  Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.  Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.

>>>>

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.  Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).  In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts depending on monitoring agency.  Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.

For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.

Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples from light microscopy that enumerated major taxa by cell concentrations and qualitative presence/absence samples.  Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.  Evaluation herein of HABs data included specific focus on the red tide organism *Karenia brevis*.  Data for *K. brevis* were obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).  Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St. Petersberg) impacted by the event. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx) that provides reports received by FWC via the state's Marine Fish Kill Hotline.

Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.  Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17]. Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.  This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant. Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.  Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.  Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.  Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.  Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.
